# Week 5: Dimensionality reduction techniques

## Overview of the data and data wrangling


Packages for the session
```{r,error=FALSE, warning=FALSE, message=FALSE}
rm(list=ls())
library(GGally)
library(ggplot2)
library(dplyr)
library(corrplot)
library(tidyr)
```


We read the data and take a look at the summary

```{r,error=FALSE, warning=FALSE, message=FALSE}
human <- read.csv("~/IODS-project/data/human.csv", sep  =",", header = T, row.names = 1)
names(human)
summary(human)
```

We have 155 observations and 8 variables from OECD's Human development report.
Link to  more information: http://hdr.undp.org/en/content/human-development-index-hdi 
Next we look graphically at the relations 

```{r,error=FALSE, warning=FALSE, message=FALSE}
my_fn <- function(data, mapping, ...){
  p <- ggplot(data = data, mapping = mapping) + 
    geom_point() + 
    geom_smooth(method=loess, fill="red", color="red", ...) +
    geom_smooth(method=lm, fill="blue", color="blue", ...)
  p
}
ggpairs(human, lower = list(continuous = my_fn))

cor(human) %>% corrplot
```

Many of the variables are not perfectly normally distributed and are skewed either to the right or left.
There is some obvious (negative) correlations for example maternal mortality ratio and life expectancy.

##Principal component analysis PCA  (Not standardized data & standardized data)

We proceed to PCA for the not standardized human data.

```{r,error=FALSE, warning=FALSE, message=FALSE}
pca_human_not_std <- prcomp(human)
sum_pca_human_not_std <- summary(pca_human_not_std)
pca_pr_not_std <- round(100*sum_pca_human_not_std$importance[2, ], digits = 3)
pca_pr_not_std
pc_lab_not_std <- paste0(names(pca_pr_not_std), " (", pca_pr_not_std, "%)")
biplot(pca_human_not_std, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab_not_std[1], ylab = pc_lab_not_std[2])
```

There is two principal components catching any variance.
The first principal component catches 99.99% of the variance.

How do things change when we standardize?

```{r,error=FALSE, warning=FALSE, message=FALSE}
human_std <- scale(human)
pca_human <- prcomp(human_std)
sum_pca_human <- summary(pca_human)
pca_pr <- round(100*sum_pca_human$importance[2, ], digits = 3)
pca_pr
```
```{r,error=FALSE, warning=FALSE, message=FALSE}
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])
```

We see that variance captured changes dramatically when the human data is standardized and for meaningful research it is necessary.
Now all eight principal components catch variance.First component catches only 53,6 % now.
X-axis is representing PC1 and y-axis is representing  PC2.
The direction and lenght of arrows represents how increase in each variable affects where the observation will be plotted.
The points where countries are is the linear combination of the "arrows". We can see that the rich countries are on the left side and poorer on the right (PC1). The vertical axis tells about countries equality.

##Tea time

For tea time we need to add factominer library. Let's read the data.

```{r,error=FALSE, warning=FALSE, message=FALSE}
library(FactoMineR)
tea <- read.table("http://factominer.free.fr/book/tea.csv",header=TRUE,sep=";")
colnames(tea)
str(tea)
summary(tea)
```

We have 300 observations from 36 variables. Next we observe variables graphically.
We include only variables: "variety", "how", "format", "sugar", "place.of.purchase", "after.lunch" for our visualization to make it clearer.

```{r,error=FALSE, warning=FALSE, message=FALSE}
tea_time <- dplyr::select(tea, variety, how, format, sugar, place.of.purchase, after.lunch)
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

We proceed to do multiple correspondence analysis and visualize it.

```{r,error=FALSE, warning=FALSE, message=FALSE}
# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), habillage = "quali")
```

Dimensions 1 (15.24 %) and 2 (14.23 %) capture almost 30% of the variance.